---
title: "Lab 5: Marginal Effects, Omitted Variables, Sensitivity Analysis, and a Brief Introduction to `regweight`"
output: html_document
date: "`r Sys.Date()`"
---

## Overview

This lab follows a simple workflow:

1. use a linear regression to describe the conditional expectation function,
2. interpret selected coefficients as marginal effects,
3. show how omitted-variable bias changes the coefficient of interest,
4. use sensitivity analysis to ask how strong an unobserved confounder would need to be to change the conclusion,
5. briefly illustrate how the `regweight` package can be used to inspect the implicit regression weights behind a coefficient.

We use the LaLonde job-training data from the `CBPS` package throughout.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r packages}
install.packages("remotes")

remotes::install_github(
  "ddimmery/regweight",
  upgrade = "never",
  dependencies = TRUE,
  build_vignettes = FALSE
)
packages <- c("dplyr", "ggplot2", "broom", "knitr", "CBPS", "sensemakr", "regweight")
missing_pkgs <- packages[!vapply(packages, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_pkgs) > 0) {
  stop(
    "Please install these packages before knitting: ",
    paste(missing_pkgs, collapse = ", ")
  )
}

invisible(lapply(packages, library, character.only = TRUE))
```

## 1. Data preparation

The outcome is earnings in 1978. The treatment is participation in the training program. We also include a standard set of pre-treatment controls.

```{r data-prep}
data("LaLonde", package = "CBPS")

df <- LaLonde %>%
  filter((exper == 1 & treat == 1) | (exper == 0 & treat == 0)) %>%
  mutate(
    log_re78 = log(re78 + 1),
    re74_log = log(re74 + 1),
    re75_log = log(re75 + 1),
    treat = as.numeric(treat)
  )

analysis_vars <- c(
  "log_re78", "treat", "age", "educ", "black", "hisp",
  "married", "nodegr", "re74_log", "re75_log", "re74.miss"
)

summary_df <- df %>%
  select(all_of(analysis_vars)) %>%
  summary()

summary_df
```

## 2. Conditional expectation function and marginal effects

Our main regression is:

\[
E[\log(re78_i + 1) \mid D_i, X_i] = \alpha + \rho D_i + X_i'\beta.
\]

Here, `treat` is binary, so the coefficient on `treat` is interpreted as the conditional difference in expected log earnings between treated and untreated units, holding the observed covariates fixed.

```{r main-model}
main_formula <- log_re78 ~ treat + age + educ + black + hisp + married +
  nodegr + re74_log + re75_log + re74.miss

m_full <- lm(main_formula, data = df)

broom::tidy(m_full) %>%
  knitr::kable(digits = 3, caption = "Main regression")
```

### Interpreting marginal effects

For a binary treatment, the marginal effect is a conditional mean difference. For a continuous regressor in a linear model, the coefficient itself is the marginal effect.

```{r marginal-effects}
treat_effect <- coef(m_full)["treat"]
age_effect <- coef(m_full)["age"]

marginal_effects <- tibble::tibble(
  quantity = c("Effect of treatment (binary D)", "Marginal effect of age (continuous X)"),
  estimate = c(treat_effect, age_effect)
)

marginal_effects %>%
  knitr::kable(digits = 3, caption = "Two coefficient interpretations in the linear model")
```

We can also show the treatment effect using predictions for the same covariate profile under `treat = 0` and `treat = 1`.

```{r prediction-example}
profile <- df %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    educ = mean(educ, na.rm = TRUE),
    black = mean(black, na.rm = TRUE),
    hisp = mean(hisp, na.rm = TRUE),
    married = mean(married, na.rm = TRUE),
    nodegr = mean(nodegr, na.rm = TRUE),
    re74_log = mean(re74_log, na.rm = TRUE),
    re75_log = mean(re75_log, na.rm = TRUE),
    re74.miss = mean(re74.miss, na.rm = TRUE)
  )

pred_data <- bind_rows(
  mutate(profile, treat = 0),
  mutate(profile, treat = 1)
)

pred_data$pred_log_re78 <- predict(m_full, newdata = pred_data)
pred_data

pred_diff <- diff(pred_data$pred_log_re78)
pred_diff
```

## 3. Omitted-variable bias

We now compare the treatment coefficient across models with different control sets.

- `m_naive`: no controls.
- `m_demo`: demographics only.
- `m_short`: adds 1974 earnings, but still omits 1975 earnings.
- `m_full`: adds 1975 earnings as well.

```{r ovb-models}
m_naive <- lm(log_re78 ~ treat, data = df)

m_demo <- lm(
  log_re78 ~ treat + age + educ + black + hisp + married + nodegr,
  data = df
)

m_short <- lm(
  log_re78 ~ treat + age + educ + black + hisp + married + nodegr +
    re74_log + re74.miss,
  data = df
)

model_compare <- bind_rows(
  broom::tidy(m_naive) %>% filter(term == "treat") %>% mutate(model = "Naive"),
  broom::tidy(m_demo) %>% filter(term == "treat") %>% mutate(model = "Demographics only"),
  broom::tidy(m_short) %>% filter(term == "treat") %>% mutate(model = "Omit re75_log"),
  broom::tidy(m_full) %>% filter(term == "treat") %>% mutate(model = "Full model")
) %>%
  select(model, estimate, std.error, statistic, p.value)

model_compare %>%
  knitr::kable(digits = 3, caption = "How the treatment coefficient changes across specifications")
```

### Exact decomposition for one omitted variable

Now we use the omitted-variable bias identity with one omitted covariate: `re75_log`.

- `delta_tilde`: effect of `re75_log` on the outcome in the long regression.
- `gamma_tilde`: residual association between `re75_log` and `treat`, conditional on the other controls.
- `delta_tilde * gamma_tilde`: implied bias from omitting `re75_log`.

```{r ovb-decomposition}
controls_z <- c("age", "educ", "black", "hisp", "married", "nodegr", "re74_log", "re74.miss")

aux_formula <- as.formula(
  paste("re75_log ~ treat +", paste(controls_z, collapse = " + "))
)

aux_model <- lm(aux_formula, data = df)

delta_tilde <- coef(m_full)["re75_log"]
gamma_tilde <- coef(aux_model)["treat"]

bias_from_formula <- delta_tilde * gamma_tilde
bias_from_models <- coef(m_short)["treat"] - coef(m_full)["treat"]

ovb_table <- tibble::tibble(
  quantity = c(
    "delta_tilde: effect of re75_log on log_re78",
    "gamma_tilde: residual imbalance of re75_log with treat",
    "delta_tilde * gamma_tilde",
    "Difference between short and full treatment coefficients"
  ),
  value = c(delta_tilde, gamma_tilde, bias_from_formula, bias_from_models)
)

ovb_table %>%
  knitr::kable(digits = 4, caption = "OVB decomposition for omitting re75_log")
```

The last two rows should be almost identical up to numerical precision. That is the linear omitted-variable-bias decomposition at work.

## 4. Partial \(R^2\) and sensitivity analysis

Sensitivity analysis asks how strong an unobserved confounder would need to be to change the treatment estimate. The two key quantities are:

- outcome-side partial \(R^2\): how much of the residual outcome variation a confounder explains,
- treatment-side partial \(R^2\): how much of the residual treatment variation it explains.

We first compute the partial \(R^2\) of an observed benchmark variable, `re75_log`, and then use those values to calibrate hypothetical confounders.

### A helper for partial \(R^2\)

```{r partial-r2-helper}
partial_r2 <- function(reduced_model, full_model) {
  r2_reduced <- summary(reduced_model)$r.squared
  r2_full <- summary(full_model)$r.squared
  (r2_full - r2_reduced) / (1 - r2_reduced)
}
```

### Benchmark partial \(R^2\) for `re75_log`

```{r benchmark-r2}
# Outcome-side partial R^2 for re75_log
m_y_reduced <- lm(
  log_re78 ~ treat + age + educ + black + hisp + married + nodegr +
    re74_log + re74.miss,
  data = df
)

m_y_full <- lm(
  log_re78 ~ treat + age + educ + black + hisp + married + nodegr +
    re74_log + re74.miss + re75_log,
  data = df
)

r2_y_benchmark <- partial_r2(m_y_reduced, m_y_full)

# Treatment-side partial R^2 for re75_log
m_d_reduced <- lm(
  treat ~ age + educ + black + hisp + married + nodegr + re74_log + re74.miss,
  data = df
)

m_d_full <- lm(
  treat ~ age + educ + black + hisp + married + nodegr + re74_log + re74.miss + re75_log,
  data = df
)

r2_d_benchmark <- partial_r2(m_d_reduced, m_d_full)

benchmark_table <- tibble::tibble(
  benchmark = "re75_log",
  outcome_side_partial_r2 = r2_y_benchmark,
  treatment_side_partial_r2 = r2_d_benchmark
)

benchmark_table %>%
  knitr::kable(digits = 4, caption = "Benchmark partial R-squared values")
```

### Manual contour plot

The graph below is built from a grid of hypothetical confounders. Each point in the graph is a pair:

\[
\left(R^2_{D \sim U \mid X},\; R^2_{Y \sim U \mid D, X}\right).
\]

For each pair we compute the implied bias using:

\[
|\text{bias}| = se(\hat\rho)\sqrt{\frac{R^2_{Y \sim U \mid D,X}R^2_{D \sim U \mid X}}{1 - R^2_{D \sim U \mid X}} \cdot df }.
\]

The red dashed contour is the set of confounder strengths that would reduce the treatment estimate to zero.

```{r sensitivity-grid}
estimate_treat <- coef(m_full)["treat"]
se_treat <- summary(m_full)$coefficients["treat", "Std. Error"]
df_treat <- df.residual(m_full)

sens_grid <- expand.grid(
  r2_treatment = seq(0.001, 0.25, length.out = 150),
  r2_outcome = seq(0.001, 0.25, length.out = 150)
) %>%
  mutate(
    implied_bias = se_treat * sqrt((r2_outcome * r2_treatment / (1 - r2_treatment)) * df_treat),
    adjusted_estimate = estimate_treat - implied_bias
  )

# Simple didactic benchmark scaling for plotting.
# This is useful for teaching the logic of the graph.
benchmark_points <- tibble::tibble(
  label = c("1x re75_log", "2x re75_log"),
  r2_treatment = pmin(c(1, 2) * r2_d_benchmark, 0.99),
  r2_outcome = pmin(c(1, 2) * r2_y_benchmark, 0.99)
) %>%
  mutate(
    implied_bias = se_treat * sqrt((r2_outcome * r2_treatment / (1 - r2_treatment)) * df_treat),
    adjusted_estimate = estimate_treat - implied_bias
  )

benchmark_points %>%
  knitr::kable(digits = 4, caption = "Benchmark-calibrated hypothetical confounders")
```

```{r sensitivity-plot, fig.width=8, fig.height=6}
ggplot(sens_grid, aes(x = r2_treatment, y = r2_outcome, z = adjusted_estimate)) +
  geom_contour(color = "grey55", breaks = pretty(sens_grid$adjusted_estimate, n = 6)) +
  geom_contour(color = "red", linetype = "dashed", linewidth = 1, breaks = 0) +
  geom_point(
    data = benchmark_points,
    aes(x = r2_treatment, y = r2_outcome),
    inherit.aes = FALSE,
    size = 2
  ) +
  geom_text(
    data = benchmark_points,
    aes(x = r2_treatment, y = r2_outcome, label = paste0(label, "\nAdj. est = ", round(adjusted_estimate, 3))),
    inherit.aes = FALSE,
    nudge_y = 0.012,
    size = 3.2
  ) +
  labs(
    title = "Sensitivity-analysis contour plot",
    x = expression(R[D %~% U %.% X]^2),
    y = expression(R[Y %~% U %.% D + X]^2),
    caption = "Grey curves: constant adjusted estimates. Red dashed curve: combinations that reduce the treatment estimate to zero."
  ) +
  theme_minimal()
```

### `sensemakr` implementation

The manual contour plot above is useful for understanding the logic. The `sensemakr` package automates a formal version of this exercise.

```{r sensemakr-example}
sens <- sensemakr::sensemakr(
  model = m_full,
  treatment = "treat",
  benchmark_covariates = "re75_log",
  kd = c(1, 2)
)

sens
```

```{r sensemakr-plot, fig.width=8, fig.height=6}
plot(sens)
```

## 5. Brief bridge to machine learning

A linear model is a simple way to describe the conditional expectation function. Machine-learning models pursue the same goal, but with more flexible functional forms, many controls, and automatic interactions or nonlinearities.

This can help with *functional-form misspecification* for observed covariates. It does **not** solve omitted-variable bias from variables that are never observed. That is why sensitivity analysis still matters even if a more flexible predictive model is used.

## 6. Brief introduction to `regweight`

The `regweight` package is useful when you want to know **which observations matter most for a particular regression coefficient**. Instead of focusing on global leverage, it focuses on the implicit weights behind one selected term.

Below we ask for the implicit regression weights behind the coefficient on `treat` in the full model.

```{r regweight-basic}
rw_treat <- regweight::calculate_weights(m_full, "treat")

hist(rw_treat)
```

We can also inspect how those implicit weights vary over a discrete covariate.

```{r regweight-discrete}
plot(rw_treat, df$married)
```

And we can compare the nominal and implicit samples for a few covariates.

```{r regweight-summary}
summary(
  rw_treat,
  dplyr::select(df, age, married, nodegr, re74_log),
  output = "html"
)
```

### Interpretation

A useful reading strategy is:

- if the weight distribution is very concentrated, a relatively small set of observations is driving the coefficient,
- if the implicit sample looks very different from the nominal sample, the regression coefficient may be learning mostly from a specific part of the data,
- this is especially useful when you worry that the OLS estimand may be giving more influence to some units than others.

## 7. Suggested discussion questions

1. How does the treatment coefficient change as we add more controls?
2. Does omitting `re75_log` produce bias in the expected direction?
3. How large would a hidden confounder need to be to drive the treatment estimate to zero?
4. Is `re75_log` a substantively strong benchmark?
5. Which observations receive more implicit weight in the `treat` coefficient according to `regweight`?
